---
title: "Manipulating, analyzing and exporting data with tidyverse"
author: "Tobin Magle"
teaching: 
exercises: 
questions:
- "What is the tidyverse? Why should I use it to manipulate data?"
objectives:
- "Describe the purpose of the **`dplyr`** and **`tidyr`** packages."
- "Select certain columns in a data frame with the **`dplyr`** function `select`."
- "Select certain rows in a data frame according to filtering conditions with the **`dplyr`** function `filter`."
- "Link the output of one **`dplyr`** function to the input of another function with the 'pipe' operator `%>%`."
- "Add new columns to a data frame that are functions of existing columns with `mutate`."
- "Use `summarize`, `group_by`, and `count` to split a data frame into groups of observations, apply summary statistics for each group, and then combine the results."
- "Describe the concept of a wide and a long table format and for which purpose those formats are useful."
- "Describe what key-value pairs are."
- "Reshape a data frame from long to wide format and back with the `spread` and `gather` commands from the **`tidyr`** package."
- "Export a data frame to a .csv file."
keypoints:
- "The tidyverse is built for data manipulation."
- "The `read_csv` function creates tibbles instead of data frames."
- "The `select` function picks particular columns based on names."
- "The `filter` function picks rows based on values."
- "The `mutate` function creates new columns based on the value of other columns."
- "The `group_by` and `summarize` functions can be used to create summary tables."
- "The `lubridate` package is useful for manipulating date data."
- "The `write_csv` function exports tibbles into a .csv file."
source: Rmd
---

```{r, echo=FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(knitr)

source("../bin/chunk-options.R")
#knitr::opts_chunk$set(message = FALSE, warnings = FALSE)

data(iris)
iris<-as_tibble(iris)

phys_data <- read_csv("data/Physiology_Environmental_Data.csv")
```

# Data Manipulation using **`dplyr`**

Bracket subsetting is handy, but it can be cumbersome and difficult to read, especially for complicated operations. **`dplyr`**. **`dplyr`** is a package for making tabular data manipulation easier.

Packages in R are basically sets of additional functions that let you do more stuff. The functions we've been using so far, like `str()` or `data.frame()`, come built into R; packages give you access to more of them. Before you use a package for the first time you need to install it on your machine, and then you should import it in every subsequent R session when you need it. 

You should already have installed the **`tidyverse`** package. This is an "umbrella-package" that installs several packages useful for data analysis which work together well such as **`tidyr`**, **`dplyr`**, **`ggplot2`**, **`tibble`**, etc.


The **`tidyverse`** package tries to address 3 common issues that arise when doing data analysis with some of the functions that come with R:

1. The results from a base R function sometimes depend on the type of data.

2. Using R expressions in a non standard way, which can be confusing for new learners.

3. Hidden arguments, having default operations that new learners are not aware of.

We have seen in our previous lesson that when building or importing a data frame, the columns that contain characters (i.e., text) are coerced (=converted) into the `factor` data type. We had to set **`stringsAsFactors`** to **`FALSE`** to avoid this hidden argument to convert our data type. 

This time we will use the **`tidyverse`** package to read the data and avoid having to set **`stringsAsFactors`** to **`FALSE`**

#### To load the tidyverse package set, including dplyr, type:

```{r load-tidy}
## load the tidyverse packages, including dplyr
library("tidyverse")
```

## What is **`dplyr`**?

The package **`dplyr`** provides easy tools for the most common data manipulation tasks like selecting columns, filtering rows, creating new columns, and creating summary tables. 

To learn more about **`dplyr`** after the workshop, you may want to check out this [handy data transformation with **`dplyr`** cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).

We'll read in our data using the `read_csv()` function, from the tidyverse package **`readr`**, instead of `read.csv()`.

```{r readcsv}
phys_data <- read_csv("data/Physiology_Environmental_Data.csv")
iris <- read_csv("data/iris.csv")

## inspect the data
str(phys_data)
str(iris)
```

To preview the data, you can click on the name of the tibble in your environment or use the `View` function. 

```{r view-data, eval=FALSE}
## preview the data
View(phys_data)
```
Either of these methods will open a window with the dataset in the same area as your script file.

#### Tibbles
Notice that the class of the data is now `tbl_df`

This is referred to as a "tibble". Tibbles tweak some of the behaviors of the data frame objects we introduced in the previous episode. The data structure is very similar to a data frame. For our purposes the only differences are that:

1. In addition to displaying the data type of each column under its name, it only prints the first few rows of data and only as many columns as fit on one screen.

2. Columns of class `character` are never converted into factors.


We're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statisitcs on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

## Selecting columns and filtering rows

To select columns of a data frame, use `select()`. The first argument
to this function is the data frame (`iris`), and the subsequent
arguments are the columns to keep.

```{r select1}
select(iris, Sepal.Length, Sepal.Width)
```


To select all columns *except* certain ones, put a "-" in front of
the variable to exclude it.

```{r select2}
select(iris, -Sepal.Length, -Sepal.Width)
```

This will select all the variables in `iris` except `Sepal.Length`
and `Sepal.Width`.

To choose rows based on a specific criteria, use `filter()`:

```{r, filter1}
filter(iris, Sepal.Length>5)
```

## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary data frame and use
that as input to the next function, like this:

```{r, pipes1}
iris2 <- filter(iris, Sepal.Length < 5)
iris_small <- select(iris2, Species, Sepal.Length, Sepal.Width)
```

This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r, pipes2}
iris_sml <- select(filter(iris, Sepal.Length < 5), Species, Sepal.Length, Sepal.Width)
```

This is handy, but can be difficult to read if too many functions are nested, as R evaluates the expression from the inside out (in this case, filtering, then selecting).

The last option, *pipes*, are a recent addition to R. Pipes let you take
the output of one function and send it directly to the next, which is useful when you need to do many things to the same dataset.  Pipes in R look like `%>%` and are made available via the **`magrittr`** package, installed automatically with **`dplyr`**. If you use RStudio, you can type the pipe with <kbd>Ctrl</kbd>
+ <kbd>Shift</kbd> + <kbd>M</kbd> if you have a PC or <kbd>Cmd</kbd> + 
<kbd>Shift</kbd> + <kbd>M</kbd> if you have a Mac.

```{r, pipes3}
iris %>%
  filter(Sepal.Length < 5) %>%
  select(Species, Sepal.Length, Sepal.Width)
```

In the above code, we use the pipe to send the `iris` dataset first through `filter()` to keep rows where `Sepal.Length` is less than 5, then through `select()` to keep only the `Sepal.Length`, `Sepal.Width`, and `species` columns. Since `%>%` takes the object on its left and passes it as the first argument to the function on its right, we don't need to explicitly include the data frame as an argument to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance, in the above example, we took the data frame `iris`, *then* we `filter`ed for rows with `Sepal.Length < 5`, *then* we `select`ed columns `species`, `Sepal.Length`, and `Sepal.Width`. The **`dplyr`** functions by themselves are somewhat simple, but by combining them into linear workflows with the pipe, we can accomplish more complex manipulations of data frames.

If we want to create a new object with this smaller version of the data, we can assign it a new name:

```{r, pipes4}
iris_sml <- iris %>%
  filter(Sepal.Length < 5) %>%
  select(Species, Sepal.Length, Sepal.Width)

iris_sml
```

Note that the final data frame is the leftmost part of this expression.

> ## Challenge 1
>
>  Using pipes, subset the `phys_data` data to include observations collected in
>  2014 and retain only the columns `Year`, `Species`, and `Photo`.
> 
> 
> > ## Solution to Challenge 1
> > ```{r, answer=TRUE, eval=FALSE, purl=FALSE}
> > phys_data %>% 
> >    filter(Year == 2014) %>%
> >    select(Year, Species, Photo)
> > ```
> {: .solution}
{: .challenge}

### Mutate

Frequently you'll want to create new columns based on the values in existing columns, for example to do unit conversions, or to find the ratio of values in two columns. For this we'll use `mutate()`.

To create a new column of weight in kg:

```{r, mutate1}
iris %>%
  mutate(Sepal.Ratio = Sepal.Length/Sepal.Width)
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, mutate2}
iris %>%
  mutate(Sepal.Ratio = Sepal.Length/Sepal.Width,
         Petal.Ratio = Petal.Length/Petal.Width)
```

> ## Challenge 2 
>  Instantaneous Water-Use Efficiency (WUE) is an efficiency measure of how much carbon is gained by photosynthesis per water lost through transpiration. A higher WUE indicates that the plant is losing less water, which is better for the plant. WUE is caluclated by dividing photosynthesis by transpiration. Often, different species and functional groups have distinct WUE.
>  Create a new data frame from the `phys_data` data that meets the following
>  criteria: 
>  * is named phys_wue
>  * contains the following columns: Species, FGroup
>  * contains a new column called wue_day (wue_day = Photo/Trmmol_day)
>  * contains only rows that have a wue_day >1
>
>  Which species and functional types have the highest WUE?
>
>  **Hint**: think about how the commands should be ordered to produce this data frame!
> 
> > ## Solution to Challenge 2
> > ```{r}
> > phys_wue <- phys_data %>%
> >     mutate(wue_day = Photo/Trmmol_day) %>%
> >     filter(wue_day > 1) %>%
> >     select(Species, Fgroup, wue_day)
> > ```
> {: .solution}
{: .challenge}

# Using dates in R

You may have noticed that we have 2 columns in the `phys_data` dataset that
represent the date when the observations were made: `Year` and `DOY` (day of year).
This is a great, compact way to represent the date: it only takes 2 columns, and if you load the data into a spreadsheet program like excel, it won't be reformatted automatically. However, it's not very human readable. (I don't know what month day 169 is off the top of my head.)

R has a package called `lubridate` that can convert to and from various date formats. Let's install and load the package:

```{r dates1, eval = FALSE}
install.packages("lubridate")
library(lubridate)
```

The as.Date function converts other data types (such as the integers in the `Year` and `DOY` columns) into date format. 

For example, we could use the following syntax to get the date for DOY = 15 and Year = 2015.
```{r dates2}
one_date<-as.Date(paste(15,2015), format="%j %Y")
one_date
class(one_date)
```

To create a column called date that holds the date information in a human readable format, we can use mutate:

```{r dates3}
phys_data<-phys_data%>%
  mutate(Date = as.Date(paste(DOY,Year), format="%j %Y"))
```

This puts the newly created column at the end. To reorder the columns, we 
can use select:

```{r dates4}
phys_data<-phys_data%>%
  select(Date, everything())
```

We can also split the date format out into its individual year/month/day components using the `month`, `day` and `year` functions:

```{r dates5}
month(one_date)
day(one_date)
year(one_date)
```

> ## Challenge 3
>  Using the phys_data, create a new data frame that contains 
>  all data plus Month and Day in the 3th and 4th column.
>
> > ## Solution to Challenge 3
> > ```{r ch3-soln}
> > phys_data <- phys_data %>%
> >     mutate(Day = day(Date), 
> >            Month = month(Date))%>%
> >     select(Date, Year, Month, Day, DOY, everything())
> > ```
> {: .solution}
{: .challenge}

# Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new data sets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from data frames.

Before using `write_csv()`, we are going to create a new folder, `data_output`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data_output`
directory, so even if the files it contains are deleted, we can always
re-generate them.

We can save it as a CSV file in a new folder called `data_output` folder.
Let's save the dataset that that we created dates in.

```{r export, eval=FALSE}
write_csv(phys_data, path = "data_output/phys_dates.csv")
```

### Split-apply-combine data analysis and the summarize() function

Many data analysis tasks can be approached using the *split-apply-combine* paradigm: split the data into groups, apply some analysis to each group, and then combine the results. **`dplyr`** makes this very easy through the use of the `group_by()` function.

#### The `summarize()` function

`group_by()` is often used together with `summarize()`, which collapses each group into a single-row summary of that group.  `group_by()` takes as arguments the column names that contain the **categorical** variables for which you want to calculate the summary statistics. So to compute the mean `Sepal.Length` by `Species`:

```{r split-apply-combine}
iris %>%
  group_by(Species) %>%
  summarize(mean_sepal_length = mean(Sepal.Length))
```

You can also group by multiple columns. For this, we'll need to use the `phys_data` dataset:

```{r split-apply-combine2}
phys_data %>%
  group_by(Year, Species) %>%
  summarize(mean_photo = mean(Photo))
```

If you want to display more data, you can use the `print()` function
at the end of your chain with the argument `n` specifying the number of rows to display:

```{r, split-apply-combine3}
phys_data %>%  
  group_by(Year, Species) %>%
  summarize(mean_photo = mean(Photo)) %>% 
  print(n = 15)
```

Once the data are grouped, you can also summarize multiple variables at the same time (and not necessarily on the same variable). For instance, we could add a column indicating the minimum photosynthesis level by year and species:

```{r split-apply-combine4}
phys_data %>%
  group_by(Year, Species) %>%
  summarize(mean_photo = mean(Photo),
            min_photo = min(Photo))
```

It is sometimes useful to rearrange the result of a query to inspect the values. For instance, we can sort on `min_photo` to put the less photosynthetic first:

```{r split-apply-combine5}
phys_data %>%
  group_by(Year, Species) %>%
  summarize(mean_photo = mean(Photo),
            min_photo = min(Photo)) %>%
  arrange(min_photo)
```

To sort in descending order, we need to add the `desc()` function. If we want to sort the results by decreasing order of minimum photosynthesis:

```{r split-apply-combine6}
phys_data %>%
  group_by(Year, Species) %>%
  summarize(mean_photo = mean(Photo),
            min_photo = min(Photo)) %>%
  arrange(desc(min_photo))
```

> ## Challenge 4
> 
> The original paper containing the iris dataset distinguished the three
> types of irises based on the sepal and petal length and width, as well
> as the ratio of length to width for sepals and petals. 
> 
> For this challenge, let's replicate their analysis. 
> * Calculate the mean sepal length and sepal width for each species
> * Calculate the mean petal lengths and petal widths for each species
> * Calculate the ratio of Sepal.Length to Sepal.Width
> * Calculate the ratio of Petal.Length to Petal.Width 
> 
> Which of these values does the best job of differentiating the setsosa
> species from the others?
> 
> > ## Solution to Challenge 4
> > 
> > ```{r}
> > iris %>% group_by(Species) %>% 
> >     summarize(meanSepal.Length = mean(Sepal.Length),
> >               meanSepal.Width = mean(Sepal.Width),
> >               Sepal.ratio=mean(Sepal.Length/Sepal.Width), 
> >               meanPetal.Length=mean(Petal.Length),
> >               meanPetal.Width=mean(Petal.Width),
> >               Petal.ratio = mean(Petal.Length/Petal.Width)
> >               )
> > 
> > ```
> > Petal ratio has the largest difference between setsosa and the other 
> > species.
> > 
> {: .solution}
{: .challenge}

#### Counting

When working with data, we often want to know the number of observations found for each factor or combination of factors. For this task, **`dplyr`** provides `count()`. For example, if we wanted to count the number of rows of data for each species, we would do:

```{r count}
iris %>%
    count(Species) 
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarizing it by counting the number of observations in that group. In other words, `iris %>% count()` is equivalent to:  

```{r count2}
iris %>%
    group_by(Species) %>%
    summarise(count = n())
```

If we wanted to count *combination of factors*, such as `Species` and `Year`, we would specify the first and the second factor as the arguments of `count()`:

```{r count3}
phys_data %>%
  count(Year, Species) 
```

With the above code, we can proceed with `arrange()` to sort the table 
according to a number of criteria so that we have a better comparison. 
For instance, we might want to arrange the table above in (i) an alphabetical order of 
the levels of the species and (ii) in descending order of the count:

```{r arrange}
phys_data %>%
  count(Year, Species) %>%
  arrange(Species, desc(n))
```

> ## Challenge 5
>
> How many individuals of each species were measured in each `Year`?
>
> > ## Solution to Challenge 5
> > ```{r, answer=TRUE, purl=FALSE}
> > phys_data %>%
> >    group_by(Year)%>%
> >    count(Species) 
> > ```
> >
> {: .solution}
{: .challenge}

> ## Challenge 6
>
> Use `group_by()` and `summarize()` to find the mean, min, and max
> photosynthesis level for each species. Also add the number of
> observations (hint: see `?n`).
>
>
> > ## Solution to Challenge 6 
> >```{r}
> > phys_data %>%
> >    group_by(Species) %>%
> >    summarize(
> >        mean_photo = mean(Photo),
> >        max_photo = max(Photo),
> >        min_photo = min(Photo),
> >        n = n()
> >    )
> >```
> {: .solution}
{: .challenge}

> ## Challenge 7
>
> Which plant had the highest photosynthetic rates in each year?
> Return the columns `Year`, `Species`, and `Photo`.
>
> > ## Solution to Challenge 7
> > ```{r, answer=TRUE, purl=FALSE}
> > phys_data %>%
> >     group_by(Year) %>%
> >     filter(Photo == max(Photo)) %>%
> >     select(Year, Species, Photo) %>%
> >     arrange(Year)
> > ```
> {: .solution}
{: .challenge}
